An Early Eocene temperature gradient modelled with geological and ecological proxies

### 2022-10-07 manuscript/manuscript.Rmd
mangroves, corals, methods

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@hollis2019]. For our analyses, this dataset was restricted to the EECO and contains four commonly used geochemical climate proxies: d180, d47, Mg/Ca and TEX86 (REFS).

Do we need a description of each?

**d18O.** Measurements of the ratio of stable isotope 18O and 16O from fossils provide

**d47.**

**Mg/Ca.**

**TEX86.**

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (~34°N–32°S), with minimum sea surface temperature (~18 °C) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef]. Nevertheless, the distribution of coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 degC N [@zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO. To do so, we extract localities inferred to be Ilerdian (early Eocene) coral reefs from the compilation, excluding coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the early Eocene. Subsequently, we use statistically derived temperature limits (minimum = 21 degC, average = 27.6 degC, maximum = 29.5 degC) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 95% of the probability density within the minimum and maximum interval. [this is not quite accurate, as the mean is not exactly between 21 and 29.5, but higher. There is only 81 % probabil ]

**Mangroves.** Mangroves are distributed throughout the tropics and subtropics today. Although many factors besides temperature determine the extent of mangrove distributions, empirical temperature limits have been established for the genera *Avicennia* and *Rhizophora*, with lower mean annual SST limits of 15.6 degC and 20.8 degC, respectively [@Quisthoudt2012]. Both *Avicennia* and members of the Rhizophoraceae family were widespread in the early Eocene, but only *Avicennia* occurred in polar latitudes [@Suan2017; @Popescu2021]. Assuming that Eocene members of these mangrove taxa had similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a temperature proxy. To translate mangrove occurrences from the compilation by @Popescu2021 to a quantitative temperature proxy, we allocate locations at which *Avicennia* pollen, but no Rhizophoraceae pollen were found, a mean temperature of 19.05 degC, between the lower *Avicennia* limit (15.6 degC) and the temperature at which *Rhizophora* commonly occurs (22.5 degC), deliberately wider than indicated by the lower *Rhizophora* limit. We set the standard deviation to 1.725 such that 95 % of the probability density are within 15.6 and 22.5 degC, placing 5% probability on the possibility that an Eocene *Avicennia*-only mangrove record occurred outside of this range. We define a 95% probability density for locations with *Avicennia* and Rhizophoraceae in a similar way. To allow for greater uncertainty at the lower temperature limit, we use the empirical 20.7 degC limit. For the upper limit, we use the maximal coral reef temperature of 29.5 degC from @kleypas1999, as the compendium of @Quisthoudt2012 does not include the warmest, equatorial regions. The mean and standard deviation of locations with *Avicennia* and Rhizophoraceae are thus set to 25.1 and 2.2, respectively.

 Rhizophoraceae commonlylimit (19.05 degC), with a standard deviation of 1.725 degC, thus placing 95 % of the probability density within that interval (see Fig. 1x). The range of temperatures in which both taxa occur in the modern besides the outlier at 20.8 degC is upwards of 22.5 degC [@Quisthoudt2012]. As the study of @Quisthoudt2012 has not sufficiently represented equatorial regions, we use the upper coral temperature limit of 29.5 degC and set the 95 % probability interval for mangroves to 22.5 - 29.5  hence occurrences of both *Avicennia* and Rhizophoraceae pollen were not used in the analysis.

@@

The palaeogeographic distribution of geochemical and ecological data was reconstructed using the @merdith2021 plate rotation model via the palaeoverse R package ver. XXX (REF). The midpoint age of the EECO (51.5 Ma) coupled with the present-day coordinates of geochemical and ecological data.

>>>>

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@hollis2019]. This compilation provides  For our analyses, this dataset was restricted to the EECO and contains four commonly used geochemical climate proxies: d180, d47, Mg/Ca and TEX86 (REFS). For a detailed description of each see [@hollis2019].

- palaeoenvironment of data (which filters were applied)
- recrystallisation (excluding most d18O)

Do we need a description of each? Not in detail, 1 paragraph in total is enough
 Measurements of the ratio of stable isotope 18O and 16O from fossils provide

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (~34°N–32°S), with minimum sea surface temperature (~18 °C) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef]. Nevertheless, the distribution of coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 degC N [@zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO. To do so, we extract localities inferred to be Ilerdian (early Eocene) coral reefs from the compilation, and that can be confidently assigned to the EECO. We exclude coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (<34 deg N). Subsequently, we use statistically derived temperature limits (minimum = 21 degC, average = 27.6 degC, maximum = 29.5 degC) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures > 29.5 degC, allowing for the possibilty that Eocene coral reefs were adapted to warmer conditions than modern-day coral reefs.

**Mangroves.** Mangroves are distributed throughout the tropics and subtropics today. While factors beside SST influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6 degC) and *Rhizophora* (20.7 degC) [@Quisthoudt2012]. Both *Avicennia* and members of the Rhizophoraceae family were widespread in the early Eocene, but only *Avicennia* occurred at polar latitudes [@Suan2017; @Popescu2021]. Assuming that Eocene members of these mangrove taxa conform to the same climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator. For this analysis, published mangrove occurrence data were taken from @Popescu2021, and converted to quantitative temperature estimates. From this data, we identify two types of pollen assemblages which we ascribe different temperature distributions:

1) *Avicennia*-only assemblages: the absence of Rhizophoraceae is indicative of temperatures being between 15.6 degC (lower temperature limt of *Avicennia*) and 20.7 (lower temperature limit of *Rhizophora*). However, a value of 22.5 degC is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature. We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05 degC and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.

2) *Avicennia* and Rhizophoraceae assemblages: the presence of both groups suggests that the locality should have a minimum temperature of 20.7 degC (lower temperature limit of *Rhizophora*). As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5 degC) as coral reef localities as mangroves are also widely distributed throughout tropical regions. Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1 and a standard deviation of 2.2, with 95% probability density within the temperature limits.

 Mangrove localities containing *Avicennia* pollen, but no Rhizophoraceae pollen, a mean temperature of 19.05 degC was ascribed. The absence of Rhizophoraceae indicates that sea surface temperature values

This value is between *Avicennia*'s lower temperature limit (15.6 degC) and the temperature at which *Rhizophora* commonly occurs (22.5 degC), deliberately wider than indicated by the lower *Rhizophora* limit.

To translate mangrove occurrences from the compilation by @Popescu2021 to a quantitative temperature proxy, we allocate locations at which *Avicennia* pollen, but no Rhizophoraceae pollen were found, a mean temperature of 19.05 degC, between the lower *Avicennia* limit (15.6 degC) and the temperature at which *Rhizophora* commonly occurs (22.5 degC), deliberately wider than indicated by the lower *Rhizophora* limit. We set the standard deviation to 1.725 such that 95 % of the probability density are within 15.6 and 22.5 degC, placing 5% probability on the possibility that an Eocene *Avicennia*-only mangrove record occurred outside of this range. We define a 95% probability density for locations with *Avicennia* and Rhizophoraceae in a similar way. To allow for greater uncertainty at the lower temperature limit, we use the empirical 20.7 degC limit. For the upper limit, we use the maximal coral reef temperature of 29.5 degC from @kleypas1999, as the compendium of @Quisthoudt2012 does not include the warmest, equatorial regions. The mean and standard deviation of locations with *Avicennia* and Rhizophoraceae are thus set to 25.1 and 2.2, respectively.

@@

The palaeogeographic distribution of geochemical and ecological data was reconstructed using the @merdith2021 plate rotation model via the palaeoverse R package ver. XXX (REF). The midpoint age of the EECO (51.5 Ma), along with the present-day coordinates of geochemical and ecological data, were used for palaeorotation.
